---
title: "Statistical Analysis and Modeling in Ecology"
subtitle: "Lecture 1 - Introduction"
author: "Thiago S. F. Silva (thiago.sf.silva@stir.ac.uk)"
date: "2019/04/07 (updated: `r Sys.Date()`)"
header-includes:
   - \usepackage{tikz}
   - \usetikzlibrary{shapes.geometric,arrows}
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---
About Me
=============

**Thiago Sanna Freire Silva:** thiago.sf.silva@stir.ac.uk

*1998 - 2002*:        Biology B.Sc. - UFRN (Brazil)

*2003 - 2004*:        Remote Sensing M.Sc. - INPE (Brazil)

*2004 - 2009*:        Geography Ph.D. - University of Victoria (Canada)

*2009 - 2013*:        ~~slave~~ Postdoc at UCSB (USA) and INPE (Brazil)

*2013 - 2019*:        Assistant professor, Dept. of Geography, UNESP Rio Claro (Brazil)

*2019 - present*:     Lecturer, Biol. and Env. Sciences, University of Stirling (UK)

**Research Interests:** global change biology, ecosystem function, community assembly, plant functional ecology, biogeochemistry, macroecology, biogeography, remote sensing, data analysis, modeling.

---
class: center, middle

# About you?

---
class: center, middle

What is Statistics?
====================

---


What is Statistics?
====================

What Statistic ** is not **:

- A perfectly objective science

- A unique and foolproof method

- An automatic decision-making system

- A solution to all scientific problems

- The salvation for poor research

<br/><br/> 

> "To consult the statistician after an experiment is finished is often
only to ask him to conduct a post mortem examination.
> I can maybe say what the experiment died of. "
> *Sir* Ronald Fisher.

---

What is Statistics?
====================

Mathematics vs. Statistics: are they the same thing?

What is Statistics for?

--

Purpose of Statistics: to estimate and quantify ** uncertainties **

Uncertainties = ** probabilities **

--

How can we quantify uncertainties about something?

- Observations...

- Repetitions...

- Experiments...
---

What is Statistics?
==================

Why do we want to quantify uncertainties about something?

- To make a **decision**

- To be confident about an **explanation**

- To make a **prediction**
---

Example
=======

"A reservoir has a 60% chance of eliminating the habitat of one aquatic species, but there is a 75% chance that it will ensure the preservation of 500 km<sup>2</sup> of primary forest habitat for the next 100 years. Should I authorize or veto the construction of this reservoir?" 

"What is the relative contribution of hydrological regime, nutrient concentration and water turbidity to the growth of aquatic plants in the Amazon?"

"How much will crop productivity increase after an addition of 10 grams of Nitrogen per m<sup>2</sup>?"   
---
class: center, middle

Statistical Modeling
=====================

---

Model
=====

.pull-left[What is a ** model **?]

.pull-right[
```{r, out.width = '100% ', fig.align ='center', echo = F}
knitr :: include_graphics ('figs/bruno.jpg')
```
]

---
class: center, middle

# $Y = \beta_0 + \beta_1 X$ ?

---

```{r, out.width = '1000% ', fig.align ='center', echo = F}
knitr :: include_graphics ('figs/landsat8.jpg')
```
---

```{r, out.width = '100% ', fig.align =' center ', echo = F}
knitr :: include_graphics ('figs/casa.jpg')
```

---
What is a model?
================

- A **simplified** representation of reality

- Describes some aspects of interest while ignoring others

<br/><br/>

> "All models are wrong. Some are useful" ** George E. P. Box **

---

### A taxonomy of models

```{r, out.width = '100% ', fig.align =' center ', echo = F}
knitr :: include_graphics ('figs/models.png')
```

**These relationships are * not * obligatory! **

---

Everything is modelling!
========================

- All of our scientific knowledge is composed of models

- Mathematical models...

- Statistical models...

- **Conceptual models**

---

Statistics is modelling
======================


```{r mean as model}
x <- c(1,3,6,8,9,12,32)
mean(x)
sd(x)
```
$\mu_x$ = $E(x)$ = `r round(mean(x),2)`

We call the mean of variable $x$ the *expectation* of $x$

**Why?**

---
```{r normplot, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 10, fig.width = 18, out.width='100%'}
library(ggplot2); library(dplyr)
plot_df <- data.frame(x = seq(-30, 60, length=150)) %>% mutate(y= dnorm(x,10.14,10.3))

ggplot(plot_df,aes(x, y)) +
    geom_line(size=1) +
    geom_vline(aes(xintercept = 10.14),color=" red", size=2) +
    xlab("X") +
    ylab("Density") +
    annotate("text", x = 15, y = .002, label = "E(x)", color="red", size = 15) +
    theme_bw(base_size = 30)
    
```


---
$E(x)$: the simplest possible model
===================================

- $E(x)$ is the most likely (i.e. the *expected*) value of $x$

- We just created a model!

- Our model simplifies $x$ by focusing only on what is the most likely outcome

- Still, it gives us *some* information about $x$, i.e. , it reduces our *uncertainty* about $x$

---
class: inverse, center, middle

# About the course


---


Course Outline
==============

This course is composed of:

- ~~Blah, blah, blah, blah~~ ** Theoretical classes **

- ~~Fighting with R~~ ** Practical exercises **

- ~~Zzzzzzzzzzzzzzz~~ ** Paper discussions **

Assessments:

- ** Exercises: ** your scripts from practical classes

- ** Mini-project: ** group statistical analysis project

    - Use your own data or any available data
---

What we will cover
=======================

- Crash course into probabilities

- Probability distributions

- Frequentist error analysis ( $\ne$ hypothesis testing :woman_facepalming:)

- Data organization and management

- Data visualization

- General inear models (regression, ANOVA, ANCOVA)

- The very basics of randomization methods

- The very basics of machine learning for data analysis and prediction
---

What we will NOT cover
======================


- GLMs and GLMMs

- SEMS, path analysis, any other "fancy method"

- Multivariate analysis (Tadeu Siqueira)

### Why?

This course is focused on making you *THINK* statistically

Once you do, you can teach yourself any method

We try to avoid [statistical machismo](https://dynamicecology.wordpress.com/2012/09/11/statistical-machismo/)


---
Why use R?
==========

.pull-left[
** R: ** a language for *statistical computing* 

- Use of R is **compulsory** for this course

- Free (as in beer), free (as in speech)

- The "standard" language for communicating ecological data analysis

- Difficult at first, but you save a lot of time in the long run

- Programming = freedom

- Everyone should install **the latest version of R** and **RStudio**.

]

.pull-right[

```{r, out.width = '100% ', fig.align ='center ', echo = F}
knitr :: include_graphics ('figs/crybaby.jpg')
```

http://www.jillgreenberg.com/
]

