---
title: "Statistical Analysis and Modeling in Ecology"
subtitle: "Lecture 1 - Introduction"
author: "Thiago S. F. Silva - thiago.sf.silva@stir.ac.uk"
date: "2019/04/07 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---
class: center, middle, inverse


# Introductions


```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_accent(
  base_color = "#2c4fa3",
  header_font_google = google_font("Francois One"),
  text_font_google   = google_font("Roboto Condensed", "400", "400i"),
  code_font_google   = google_font("Fira Code"),
  text_font_size = '1.5em',
  title_slide_background_image = 'https://www.ox.ac.uk/sites/files/oxford/styles/ow_large_feature/public/field/field_image_main/Mathematics%20and%20Statistics.jpg',
  link_color = "#eb0037" 
        
)
```
---

About Me
=============
<small>
**Thiago Sanna Freire Silva:** thiago.sf.silva@stir.ac.uk

*1998 - 2002*:        Biology B.Sc. - UFRN (Brazil)

*2003 - 2004*:        Remote Sensing M.Sc. - INPE (Brazil)

*2004 - 2009*:        Geography Ph.D. - University of Victoria (Canada)

*2009 - 2013*:        ~~slave~~ Postdoc at UCSB (USA) and INPE (Brazil)

*2013 - 2019*:        Assistant professor, Dept. of Geography, UNESP Rio Claro (Brazil)

*2019 - present*:     Lecturer, Biol. and Env. Sciences, University of Stirling (UK)

**Research Interests:** global change biology, ecosystem function, community assembly, plant functional ecology, biogeochemistry, macroecology, biogeography, remote sensing, data analysis, modeling.
</small>
---
class: center, middle

# Who are you?

---
class: center, middle, inverse

What is Statistics?
====================

---


What is Statistics?
====================

What Statistics ** is not **:

- A completely objective science

- A unique and foolproof method

- An automatic decision-making system

- A solution to all scientific problems

- The salvation for poor research


> *"To consult the statistician after an experiment is finished is often
only to ask him to conduct a post mortem examination.
> I can maybe say what the experiment died of. "*
> **- Ronald Fisher**.

---

What is Statistics?
====================

Mathematics vs. Statistics: are they the same thing?

--

What is Statistics for?

--

Purpose of Statistics: to estimate and quantify ** uncertainties **

Uncertainties = ** probabilities **

How can we quantify uncertainties about something?

--

- Observations...

- Repetitions...

- Experiments...
---

What is Statistics?
==================

Why would we want to quantify uncertainties about something?
--

- To make a **decision**

- To be confident about an **explanation**

- To make a **prediction**
---

Examples
=======

"A reservoir has a 60% chance of eliminating the habitat of one aquatic species, but there is a 75% chance that it will ensure the preservation of 500 km<sup>2</sup> of primary forest habitat for the next 100 years. Should I authorize or veto the construction of this reservoir?" 

"What is the relative contribution of hydrological regime, nutrient concentration and water turbidity to the growth of aquatic plants in the Amazon?"

"How much will crop productivity increase after an addition of 10 grams of Nitrogen per m<sup>2</sup> ?"

---
What statistics is
==================

- A method for quantifying uncertainty

- A method to **help** us make decisions

- A method to **help** us make interpretations

- A method to **help** us make make predictions

**Good computer:** R$ 10,000.00

**PhD Student:**   R$ 2.200,00 $\times$ 48 months = R$ 105,600

**Don't make yourself useless**!

---
class: center, middle, inverse

Statistical Modeling
=====================

---

Model
=====

.pull-left[What is a ** model **?]

.pull-right[
```{r, out.width = '150% ', fig.align ='center', echo = F}
knitr :: include_graphics ('figs/bruno.jpg')
```
]

---
class: center, middle

# $Y = \beta_0 + \beta_1 X$

---

```{r, out.width = '1000% ', fig.align ='center', echo = F}
knitr :: include_graphics ('figs/landsat8.jpg')
```
---

```{r, out.width = '100% ', fig.align =' center ', echo = F}
knitr :: include_graphics ('figs/casa.jpg')
```

---
What is a model?
================

- A **simplified** representation of reality

- Describes some aspects of interest while ignoring others

- Doesn't have to be expressed using math!

<br/><br/>

> "All models are wrong. Some are useful" 
> **- George E. P. Box **

---

### A taxonomy of models

```{r, fig.width = 20, out.height='100%', fig.align='center', echo = F}

DiagrammeR::grViz("
digraph graph2 {

graph [layout = dot]

# node definitions with substituted label text
node [shape = rectangle, width = 4, fillcolor = Biege]
a [label = '@@1']
b [label = '@@2']
c [label = '@@3']
d [label = '@@4']
e [label = '@@5']
f [label = '@@6']
g [label = '@@7']
h [label = '@@8']
i [label = '@@9']

a -> b 
a -> c
c -> d
c -> e
d -> f -> h
e -> g -> i

}

[1]: 'Models' 
[2]: 'Conceptual' 
[3]: 'Quantitative' 
[4]: 'Mathematical' 
[5]: 'Statistical' 
[6]: 'Deterministic' 
[7]: 'Stochastic' 
[8]: 'Theoretical' 
[9]: 'Empirical' 

",width=800,height=400)

```

**These relationships are * not * obligatory! **

---

Everything is modelling!
========================

- All of our scientific knowledge is composed of models

- Mathematical models...

- Statistical models...

- **Conceptual models**

---

Statistics is modelling
======================


```{r mean as model, tidy=TRUE}
x <- c(1,3,6,8,9,12,32)
mean(x)
```

<br/>

$\mu_x$ = $E(x)$ = `r round(mean(x),2)`

<br/>


We call the mean of variable $x$ the *expectation* of $x$

**Why?**

---
```{r normplot, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 10, fig.width = 18, out.width='100%'}
library(ggplot2); library(dplyr)
plot_df <- data.frame(x = seq(-30, 60, length=150)) %>% mutate(y= dnorm(x,10.14,10.3))

ggplot(plot_df,aes(x, y)) +
    geom_line(size=1) +
    geom_vline(aes(xintercept = 10.14),color=" red", size=2) +
    xlab("X") +
    ylab("Density") +
    annotate("text", x = 15, y = .002, label = "E(x)", color="red", size = 15) +
    theme_bw(base_size = 30)
    
```


---
$E(x)$: the simplest model
===================================

- $E(x)$ is the most likely (i.e. the *expected*) value of $x$

- We just created a model!

- Our model simplifies $x$ by focusing on what is the most likely outcome

- Still, it gives us *some* information about $x$, i.e. , it reduces our *uncertainty* about $x$


---

## Going beyond: improving the models


From here on, we can see every statistical analysis as:

- Defining on a model form

- Fitting the model to data

- Assessing the confidence we have about the model fit

- Comparing models to decide which one is best represents the data

---

```{r normplot2, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 10, fig.width = 18, out.width='100%'}
plot_df <- data.frame(x = seq(-30, 60, length=150)) %>%
    mutate(y= dnorm(x,10.14,10.3),y2= dnorm(x,10.14,5))

ggplot(plot_df,aes(x, y)) +
    geom_line(size=1) +
    geom_line(aes(x=x,y=y2), size=1, color = " Blue") +
    geom_vline(aes(xintercept = 10.14),color=" red", size=2) +
    xlab("E(X)") +
    ylab("Density") +
    #annotate("text", x = 15, y = .002, label = "E(x)", color="red", size = 15) +
    theme_bw(base_size = 30)
```

Where do we have more *uncertainty*?

---
class: inverse, center, middle

# About the course


---
Course Outline
==============

This course is composed of:

- ~~Blah, blah, blah, blah~~ ** Theoretical classes **

- ~~Fighting with R~~ ** Practical exercises **

- ~~Zzzzzzzzzzzzzzz~~ ** Paper discussions **

Assessments:

- ** Exercises: ** your scripts from practical classes

- ** Mini-project: ** group statistical analysis project

    - Use your own data or any available data
---

What we will cover
=======================

- Crash course into probabilities

- Probability distributions

- Frequentist error analysis ( $\ne$ hypothesis testing :woman_facepalming:)

- Data organization and management

- Data visualization

- General inear models (regression, ANOVA, ANCOVA)

- The very basics of randomization methods

- The very basics of machine learning for data analysis and prediction
---

What we will NOT cover
======================


- GLMs and GLMMs

- SEMS, path analysis, any other "fancy method"

- Multivariate analysis (Tadeu Siqueira)

### Why?

This course is focused on making you *THINK* statistically

Once you do, you can teach yourself any method

We try to avoid [statistical machismo](https://dynamicecology.wordpress.com/2012/09/11/statistical-machismo/)


---
Why use R?
==========

.pull-left[
** R: ** a language for *statistical computing* 

- Use of R is **compulsory** for this course

- Free 

- The "standard" language for communicating ecological data analysis

- Difficult at first, but you will thank yourself in the future

]

.pull-right[

```{r, out.width = '100% ', fig.align ='center ', echo = F}
knitr :: include_graphics ('figs/crybaby.jpg')
```

http://www.jillgreenberg.com/
]

---
Why use R?
==========

.pull-left[
** R: ** a language for *statistical computing* 

- Programming = freedom

- Not different from learning lab or field protocols

- Everyone should install **the latest version of R** and **RStudio**.

]

.pull-right[

```{r, out.width = '100% ', fig.align ='center ', echo = F}
knitr :: include_graphics ('figs/crybaby.jpg')
```

http://www.jillgreenberg.com/
]
---
Summary
=======

- Statitics is **philosophy of science** expressed with numbers

- We need statistics when we have **uncertainty**

- We use statistics to **help** us:
    
    - Make a decision
    - Explain a process
    - Predict a quantity
    
- Knowledge is built using models (not all mathematical)

- All statistical analyses can be seen as models

- Our task is to find a good model to describe the data and quantify the uncertainty

---

Next steps
==========

Watch the recorded lectures on:

- Probabilities
- Hypothesis testing
- Type I and type II errors

Read the paper:

*Gigerenzer, G. (2004) Mindless Statistics. The Journal of Socio-Economics 33: 587–606.*

We meet tomorrow (09/04) at 3pm to answer questions and discuss the paper.
---
class: center, middle, inverse

```{r, out.width = '100% ', fig.align =' center ', echo = F}
knitr :: include_graphics ('https://i.stack.imgur.com/c88K3.png')
```

